---
title: "lab 8"
author: "Janelle Christensen"
date: "3/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(palmerpenguins)

# Packages for cluster analysis:
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)

```

# Intro to Cluster analysis (k-means, heirarchcal)

## Part 1. K-means clustering:

### Exploratory visualization

First map species onto point color, does it look like there is an opportunity to cluster by species?

```{r}

# Bill lenght v. depth 

ggplot(penguins) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange", "cyan4", "darkmagenta"))

```

```{r}

# fliper length v body mass
ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange","cyan4","darkmagenta"))

```
```{r}

number_est <- NbClust(penguins[3:6], min.nc = 2, max.nc = 10, method = "kmeans")

```

```{r}
number_est

```
### Create a complete, scaled version of the data

```{r}

# drop rows where any of the four size measurements are missing

penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

# only keep the colums for the four size measurements then scale them

penguins_scale <- penguins_complete %>% 
  select(ends_with("mm"), body_mass_g) %>% 
  scale()

```

### Run k-means

```{r}

penguins_km <- kmeans(penguins_scale, 3) # specify 3 groups

penguins_km$size # how many observations assigned to each cluster

```
```{r}
penguins_km$cluster # what cluster each observation in penguins_scale is assigned to
```
```{r}
# bind the cluster number to the original data used for clustering, so that we can see what cluster each penguin is assigne to

penguins_cl <- data.frame(penguins_complete, cluster_no = factor(penguins_km$cluster))

# plot flipper length v. body mass, indicating which cluster each penguin is assigne to (but also showing the actual species):

ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = cluster_no,
                 shape = species))
```
```{r}

ggplot(penguins_cl) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = cluster_no,
                 shape = species))

```

We can see subjectively that a lot of Adelies are in cluster 2, chinstaps in culuster 1 and gentoos in cluster 3, but what are the actual counts?

```{r}

penguins_cl %>% 
  count(species, cluster_no) %>% 
  pivot_wider(names_from = cluster_no, values_from = n) %>% 
  rename('Cluster 1' = '1', 'Cluster 2' = '2', 'Cluster 3' = '3')

```
## Part 2: Cluster analysis, hierachical

### Read in the data and simplify

```{r}

wb_env <- read_csv("wb_env.csv")

```
```{r}
wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20)
```

### scale the data
```{r}
# scale the nummeric variables (columns 3:7)

wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% 
  scale()

# update to add romnames

rownames(wb_scaled) <- wb_ghg_20$name

```
### Find the Euclidean distances
```{r}
euc_distance <- dist(wb_scaled, method = "euclidean")

View(euc_distance)

```

### Perform hierarchical clustering by complete linkage with `stats::hclust()`

```{r}

hc_complete <- hclust(euc_distance, method = "complete")

plot(hc_complete,  cex = 0.6, hang = -1)

```
## Now let's do it by single linkage and compare

```{r}
hc_single <- hclust(euc_distance, method = "single")

plot(hc_single, cex = 0.6, hang = -1)
```

### Make a tanglegram to compare dendrograms
```{r}

dend_complete <- as.dendrogram(hc_complete)

dend_simple <- as.dendrogram(hc_single)

```

```{r}
tanglegram(dend_complete, dend_simple)
```
```{r}
ggdendrogram(hc_complete, rotate = TRUE) +
  theme_minimal() +
  labs(x = "Country")
```

